---
title: "Flight_Paths"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Plotting Flight Paths

Here is the guide to plot the paths of flights on a map to visualize the flight traffic. All that you need is the latitude and longitude of the origin and destination endpoints.

There are various packages available but in this tutorial plotly is used.

```{r}
library(plotly)

airports <- read.csv('https://raw.githubusercontent.com/plotly/datasets/master/2011_february_us_airport_traffic.csv')

flights <- read.csv('https://raw.githubusercontent.com/plotly/datasets/master/2011_february_aa_flight_paths.csv')
head(flights)

# create a template of how the map looks like

geo <- list(
  scope = 'north america',
  projection = list(type = 'azimuthal equal area'),
  showland = TRUE,
  landcolor = toRGB("gray95"),
  countrycolor = toRGB("gray80")
)

p <- plot_geo(locationmode = 'USA-states', color = I("red")) %>%
  
  # mark various airports on the map using the latitude, longitude and airport names found in airports dataset
  
  add_markers(
    data = airports, x = ~long, y = ~lat, text = ~airport,
    size = ~cnt, hoverinfo = "text", alpha = 0.5
  ) %>%
  
  # From the flights data, plot the paths between origin and destination using their latitudes and longitudes
  
  add_segments(
    data = flights,
    x = ~start_lon, xend = ~end_lon,
    y = ~start_lat, yend = ~end_lat,
    alpha = 0.3, size = I(1), hoverinfo = "none"
  ) %>%
  
  # plot the flight paths on the layout north america as specified in the object geo
  layout(
    title = 'Feb. 2011 American Airline flight paths<br>(Hover for airport names)',
    geo = geo, showlegend = FALSE, height=800
  )
p
```



In the map, the lines going out of north america indicate that the destinations are other countries which we are not showing in this map.

This is just an introduction and you can find more in the references below.

## References

https://www.r-bloggers.com/create-air-travel-route-maps-in-ggplot-a-visual-travel-diary/

http://strimas.com/spatial/long-flights/

https://www.gis-blog.com/flight-connection-map-with-r/