
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("ggridges")
library("tidyverse")
library("cluster")
```

# (PART) Documentation (Green) {-}

# Chart : Ridgeline Plots

## Overview

This section covers how to make ridgeline plots.

##tl;dr

I want a nice example and I want it NOW!

Here's a look at the dose of theophylline administered orally to the subject on which the concentration of theophylline is obsereved:

Here is the code:

```{r, fig.width=12}
Theoph_data <- Theoph
ggplot(Theoph_data, aes(x=Dose,y=Subject,fill=Subject))+
  geom_density_ridges_gradient(scale=4) + theme_ridges()+
  scale_y_discrete(expand = c(0.01, 0)) +
  scale_x_continuous(expand = c(0.01, 0))+
  labs(x="Dose of theophylline(mg/kg)",y="Subject of experimentation")+
  ggtitle("Density estimation of dosage given to various subjects under observation")+
  theme(plot.title = element_text(hjust = 0.5))
```

For more info on this dataset, type ?datasets::Theoph into the console.

## Simple examples

Okay..much simpler please.

Let's use the 'Orange' dataset from the datasets package

```{r}
library("datasets")
head(Orange, n=5)
```

## Ridgeline Plots using ggridge

```{r}
ggplot(Orange, aes(x=circumference,y=Tree,fill = Tree))+
  geom_density_ridges(scale = 2, alpha=0.5) + theme_ridges()+
  scale_fill_brewer(palette = 4)+
  scale_y_discrete(expand = c(0.8, 0)) +
  scale_x_continuous(expand = c(0.01, 0))+
  labs(x="Circumference at Breast Height", y="Tree with ordering of max diameter")+
  ggtitle("Density estimation of circumference of different types of Trees")+
  theme(plot.title = element_text(hjust = 0.5))
```

ggridge uses two main geoms: "geom_density_ridges" and "geom_ridgeline" to plot the ridgeline density plots. They are used to plot the densities of various factors of a categorical variable on the y-axis to see its distribution over a continuous scale.

One more example:

```{r}
OrchardSprays_data <- OrchardSprays
ggplot(OrchardSprays_data, aes(x=decrease,y=treatment,fill=treatment))+
  geom_density_ridges_gradient(scale=3) + theme_ridges()+
  scale_y_discrete(expand = c(0.3, 0)) +
  scale_x_continuous(expand = c(0.01, 0))+
  labs(x="Response in repelling honeybees",y="Treatment")+
  ggtitle("Density estimation of response by honeybees to a treatment")+
  theme(plot.title = element_text(hjust = 0.5))
```

## When to Use

Ridgeline plots can be used when a number of data segments have to be plotted on a same horizontal scale. It is presented with slight overlap. Ridgeline plots are very useful to visualize the distribution of a categorical variable over time or space.

A good example where ridgeline plots will be a great example is visualizing the distribution of salary over different departments in a company.

## Considerations

The overlapping of the density plot can be controlled by adjusting the value of scale. Scale defines how much the peak of the lower curve touches the curve above.

The variable in the y-axis should be ordered(if it's a nominal variable) for better visualization of the density.

Ridgeline plots can also be used to plot histograms on the common horizontal axis rather than density plots. But doing that may not give us any valuable results.
