---
title: "Untitled"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This section describes how to get multiple independent plots into one window so that it is easier to compare and contrast between graphs. Many a times, we want to draw comparisons and it would be helpful to have all the plots on a single window/page to help draw conclusions

The packages that help us achieve this are
1) ggpubr
2) cowplot
3) par
4) gridExtra
5) grid

In this tutorial, I will use one or two packages from the above to demonstrate the idea. Each package has its own capabilities.

## Installing packages

Install package "ggpubr" and load it.

```{r}
install.packages("ggpubr")
library(ggpubr)
```

## Multiple plots on one window

In order to get multiple independent plots on the same window, use ggarrange function as follows.

```{r}
qq <- ggplot(beaver1,aes(sample=temp))+geom_qq_line()+geom_qq()+ggtitle("qq plot beaver1 temp")

hist <- ggplot(beaver1,aes(x=temp))+geom_histogram(color="black",fill="lightblue")+ggtitle("Histogram of the temperatures")

box <- ggplot(beaver1,aes(x=temp,y=time))+geom_boxplot()+ggtitle("boxplot of temperatures")

scatter <- ggscatter(beaver1,x="temp",y="time")+ggtitle("scatterplot of temperature with time")

ggarrange(qq,hist,box,scatter,ncol=2,nrow=2,widths = c(1, 1.25))
```


We can create and arrange how many ever plots we want on a single window. The function takes in a list of plots as argument and allows us to alter the width of each column in the plot. In the above code, widths=c(1,1.25) would make the second column is 1.25 times as wide as the first column. There are a lot of other options available in this function to improve the aesthetics of the graph. Please go through them to make your plot look best.


Here is another example where you want to visualize two different types of plots for the same data in one window.

```{r}
bp <- ggboxplot(beaver1,x="day",y="temp",color="day")+ggtitle("Boxplot of temperature on each day")
dp <- ggdotplot(beaver1,x="day",y="temp",color="day")+ggtitle("dot plot of temperature on each day")
ggarrange(bp,dp,ncol=2,nrow=1)
```

## Cowplot package

The cowplot package helps us to specify locations for our individual plots and also the size with which they should be plotted.

ggdraw() function gives the layout of the window. The height and width of the window can be set using arguments xlim and ylim. Default is "x" axis ranges from 0 to 1 and "y" axis ranges from 0 to 1. 

COnsidering (0.5,0.5) as the center of the window, the following code plots the graphs in the specified coordinates in the window.

```{r}
install.packages("cowplot")
library(cowplot)

ggdraw()+draw_plot(dp, x = 0, y = 0, width = 1, height = 0.5)+draw_plot(bp,x=0.5,y=0.5,width = 0.5,height = 0.5)+draw_plot(qq,x=0,y=0.5,width = 0.5,height = 0.5)
```

A lot more amazing visualizations are available in the following referencec links. I recommend taking a look at them.

## References

http://www.sthda.com/english/articles/24-ggpubr-publication-ready-plots/81-ggplot2-easy-way-to-mix-multiple-graphs-on-the-same-page/

https://cran.r-project.org/web/packages/gridExtra/vignettes/arrangeGrob.html
