#  TidyVerse

Zhisen Cai

Sometimes we may forget the way to use Tidyverse. Therefore, I create this for help. People can finish most of the basic part of tidyverse by using this. I separate it into 4 parts, manipulation between two datasets, manipulation on rows, manipulation on columns and group_by.  With the examples of each function, people can know how to use the function and how the function works.

```{r}
library(tidyverse)
```
Magrittr use pipeline to convey the data. (%>%) We have to use a dot to represent the thing we convey if it is not used in the first parameter. 
```{r}
# we can omit '.'
c(1, 3, 4, 5, NA) %>% mean(., na.rm=TRUE)
c(1, 3, 4, 5, NA) %>% mean(na.rm = TRUE)  
# we can not omit '.' in the second parameter
c(1, 3, 4, 5) %>% plot(., main=paste(., collapse=", "))
c(1, 3, 4, 5) %>% plot(main=paste(., collapse=", "))
```

### 0. Connection Bewtween two datasets
Efficiently bind multiple data frames by row and column 
bind_rows() and bind_cols() return the same type as the first input, either a data frame, tbl_df, or grouped_df
code:  bind_rows(..., .id = NULL)   bind_cols(...)
```{r}
bind_rows(
+   sample_n(iris, 10),
+   sample_n(iris, 10),
+   sample_n(iris, 10)) %>% 
  glimpse()
```
```{r}
bind_cols(
  sample_n(iris, 10),
  sample_n(iris, 10),
  sample_n(iris, 10)
) %>% glimpse()
```

### 1.combime two datasets by value: join

```{r}
# prepare data 
superheroes <- tribble(
  ~name,      ~alignment,  ~gender,   ~publisher,
  "Magneto",  "bad",       "male",    "Marvel",
  "Storm",    "good",      "female",  "Marvel",
  "Mystique", "bad",       "female",  "Marvel",
  "Batman",   "good",      "male",    "DC",
  "Joker",    "bad",       "male",    "DC",
  "Catwoman", "bad",       "female",  "DC",
  "Hellboy",  "good",      "male",    "Dark Horse Comics"
)
publishers <- tribble(
  ~publisher, ~yr_founded,
  "DC",       1934L,
  "Marvel",   1939L,
  "Image",    1992L
)
```

left_join
A left join in R is a merge operation between two data frames where the merge returns all of the rows from one table (the left side) and any matching rows from the second table. A left join in R will NOT return values of the second table which do not already exist in the first table.
left_join(a_tibble, another_tibble, by = ...)
```{r}
superheroes %>% 
  left_join(publishers, by="publisher")
```
inner join:
he inner_join keyword selects records that have matching values in both tables.It will not return unmatched rows. Therefore, we do not see the last row shown in the left_join.
code: inner_join(a_tibble, another_tibble, by = ...)
```{r}
inner_join(superheroes, publishers, by="publisher")
```
anti join:
An anti join returns the rows of the first table where it cannot find a match in the second table.
```{r}
superheroes %>%
  anti_join(publishers, by="publisher")
```

semi join
Semi joins are the opposite of anti joins: an anti-anti join, if you like.
A semi join returns the rows of the first table where it can find a match in the second table. 
code: semi_join(a_tibble, another_tibble, by = ...)
```{r}
superheroes %>%
  semi_join(publishers, by="publisher")
```
### 2. manipulation on column
we use select to choose column.
we can use helper function to choose column
starts_with() : start with a literal string
ends_with() : end with a literal string
matches(): match a regular expression
num_range(): are part of a numercial range
everything(): selects all columns
```{r}
colnames(iris)
```
```{r}
iris <- iris %>% as_tibble() 
iris %>%
  select(starts_with("Petal")) %>%
  head(10)
```
```{r}
iris %>%
  select(ends_with("Length"))%>%
  head(10)
```

```{r}
iris %>%
  select(contains(".")) %>%
  head(10)
```

select_if 
we can use select_if to choose column with logical condition or any function return true or false.
```{r}
starwars %>%
  select_if(is.numeric) %>%
  head(10)
```

```{r}
less_than_500 <- function(x) {
  sum(x) < 500
}
iris[1:4] %>%
  select_if(less_than_500)%>%
  head(10)
```

!!!! we can use ~ to represent anonymous function
```{r}
iris[1:4] %>%
  select_if(~ sum(.) < 500)%>%
  head(10)
```

We can select the data by deleting columns using minus sign.
```{r}
iris %>%
  select(-Species, -Petal.Length) %>%
  head(10)
```

```{r}
iris %>%   
  select(Species, everything(), -ends_with("Length")) %>%
  head(10)
```

rename the column
```{r}
iris %>%
  rename(sep_len=Sepal.Length, sep_wid=Sepal.Width) %>%
  names()
```

### 4. manipulation on row
```{r}
# prepare data
set.seed(896)
sw_dup <-
  starwars %>%
  select(-(films:starships)) %>%  
  sample_n(100, replace=TRUE)
head(sw_dup,10) 
```

Use arrange() to reorder the rows. 
```{r}
sw_dup %>%
  arrange(name, gender) %>%
  head(10)
```

using desc() for decreasing order
```{r}
sw_dup %>%
  arrange(desc(mass)) %>%
  head(10)
```
use distinct() to remove the duplicated rows
```{r}
sw_dup %>%
  distinct() %>%
  glimpse() %>%
  anyDuplicated()  
```

use drop_na() to remove all the rows including NA
```{r}
sw_dup %>%
  drop_na() %>%
  glimpse() %>%
  anyNA()
```
or can only drop na with specific rows
```{r}
sw_dup %>%
  drop_na(gender:species) %>%
  glimpse() %>%
  anyNA()
```

use filter to select rows
```{r}
sw_dup %>%
  filter(species == "Human", (is.na(mass) | height > 180)) %>%
  head(10)
```

use filter_all() to choose rows in all the columns
```{r}
iris[,1:4] %>%
  as_tibble() %>%  
  filter_all(any_vars(. > 7.5))
```

use filter_if() to choose rows 
```{r}
sw_dup %>% 
  filter_if(is.character, any_vars(is.na(.)))
```

### 5. build new column or mutate column
use mutate to create rows or columns
```{r}
iris %>%
  as_tibble(iris) %>%
  mutate(add_all = Sepal.Length + Sepal.Width + Petal.Length + Petal.Width) %>%
  head(10)
```

```{r}
iris %>%
  as_tibble() %>%
  mutate(median_petal_length = median(Petal.Length),
         has_long_petals = Petal.Length > median_petal_length,
         has_long_petals = as.numeric(has_long_petals)) %>%  
  select(Petal.Length, median_petal_length, has_long_petals)
```

use mutate_all to apply function to all the rows and columns
```{r}
msleep %>%
  mutate_all(tolower)
```

use mutate_if to apply function to specific row and column which satisfied some conditions
```{r}
iris %>%
  mutate_if(is.double, as.integer) %>%
  head(10)
```

use mutate_at to apply function to specific row and column
```{r}
iris %>%
  as_tibble() %>%
  mutate_at(vars(contains("Length"), contains("Width")), ~ .*10) %>%
  head(10)
```

We can use if_else() and case_when() to decide how to change the value
if_else(condition, if true:---, if false: ----)
```{r}
warpbreaks %>%
  mutate(breed = if_else(wool == "A",
                         true = "Merino",
                         false = "Corriedale")) %>%
  sample_frac(size = 0.15)
```
case_when (condition1 ~ class1,
           condition2 ~ class2,
           condition3 ~ class3,
           ......
  )
```{r}
warpbreaks %>%
  mutate(tension=case_when(tension == "H" ~ "High",
                           tension == "M" ~ "Medium",
                           tension == "L" ~ "Low",
                           TRUE ~ NA_character_)) %>%
  sample_frac(size = 0.15)
```

### 6. Group
use group_by() to create groups and use ungroup to separate group
use summarize() to find out the properties of the group
n() :count the frequency
sum(var),max(var),min(var),mean(var),median(var),sd(var)
```{r}
iris %>%
  group_by(Species) %>%
  summarise(n = n(),
            avg_len = mean(Petal.Length, na.rm=TRUE),
            med_len = median(Petal.Length))
```

We can also use summarize_if,summarize_all
```{r}
dat <- iris %>%
  group_by(Species) %>%
  summarise_if(is.numeric, list(~mean(.), ~median(.), ~min(.), ~max(.)))
dat
```